---
title: "index"
format: html
---

# Introduction

Marine aquaculture represents a critical opportunity for sustainable food production as global seafood demand continues to rise. Unlike land-based meat production, marine aquaculture can provide protein with lower environmental costs when strategically located in suitable environments (Hall et al., 2011). Recent research has demonstrated that global seafood demand could theoretically be met using less than 0.015% of the global ocean area (Gentry et al., 2017), highlighting the importance of identifying optimal locations for development.

This analysis evaluates the potential for marine aquaculture along the US West Coast by identifying Exclusive Economic Zones (EEZs) with suitable environmental conditions for oyster and black abalone cultivation. Using geospatial analysis of sea surface temperature and bathymetry data, we determine which regions offer the greatest area of suitable habitat for each species. The analysis addresses three key questions:

1. Which West Coast EEZs have the largest areas of suitable conditions for oyster aquaculture?
2. Which regions are most suitable for black abalone cultivation?
3. How do suitability patterns differ between these two species?

By developing a generalizable workflow, this analysis can be applied to assess aquaculture potential for additional species, supporting evidence-based decision-making for sustainable marine resource development.

# Setup

```{r setup}
#| message: false
#| warning: false

# Load required packages
library(tidyverse)
library(sf)
library(terra)
library(here)
library(tmap)
library(kableExtra)
library(testthat)
```

# Data Loading

## Load Exclusive Economic Zones (EEZ)

```{r load-eez}
#| message: false
#| warning: false

# Load West Coast EEZ shapefile
wc_df <- st_read("data/wc_regions_clean.shp")

# Check CRS
st_crs(wc_df)$epsg

```

## Load Sea Surface Temperature (SST) Data

```{r load-sst}
#| message: false
#| warning: false

# Load SST rasters for 2008-2012

# Define a function to read rast
read_tif <- function(year) {
  rast(file.path("data", paste0("average_annual_sst_", year, ".tif")))
}

# Apply function
sst_list <- lapply(2008:2012, read_tif)

# Combine into raster stack 
sst_stack <- rast(sst_list)

# Inspect properties
names(sst_stack)
st_crs(sst_stack["average_annual_sst_2008"])
summary(values(sst_stack))
global(sst_stack, fun = "isNA")

```

## Load Bathymetry (Depth) Data

```{r load-depth}
#| message: false
#| warning: false

# Load depth raster
depth <- rast(("data/depth.tif"))

# Check properties
st_crs(depth)
```

## Verify Coordinate Reference Systems

```{r check-crs}
# Check sst crs
crs(sst_stack, describe = TRUE)$code

# Set epsg explicitly from metadata
crs(sst_stack) <- "EPSG:4326"

# Check again
crs(sst_stack, describe = TRUE)$code

# Check extrent makes sense for this CRS
ext(sst_stack) 

# Verify sst and depth crs are the same
crs(sst_stack) == crs(depth)

# Check if wc_df crs matches
crs(sst_stack, describe = TRUE)$code == st_crs(wc_df)$epsg


```

# Data Processing

## Calculate Mean SST (2008-2012)

```{r mean-sst}
# Calculate mean SST across all years using app()
sst_mean <- app(sst_stack, mean, na.rm = TRUE)

# Visualize with plot()
plot(sst_mean)
```

## Convert SST from Kelvin to Celsius

```{r convert-temp}
# Convert from Kelvin to Celsius (subtract 273.15)
sst_mean_c <- sst_mean - 273.15

# Check temperature range with summary()
summary(sst_mean_c)
```

## Align Depth Data with SST

```{r align-depth}
# Crop depth to SST extent using crop()
depth_crop <- crop(depth, sst_mean_c)

# Resample depth to match SST resolution using resample() with method = "near"
depth_resampled <- resample(depth_crop, sst_mean_c, method = "near")

# Verify alignment 
res(depth_resampled) == res(sst_mean_c)
ext(depth_resampled) == ext(sst_mean_c)
crs(depth_resampled) == crs(sst_mean_c)
```

# Species Suitability Analysis

## Oyster Suitability

### Define Oyster Requirements

Oysters require:
- Sea surface temperature: 11-30°C
- Depth: 0-70 meters below sea level

```{r oyster-suitability}
# Reclassify SST for oysters using classify()
# Create rcl matrix: <11°C = 0, 11-30°C = 1, >30°C = 0
sst_rcl_mat_oyster <- matrix(c(-Inf, 11, 0,
                                11, 30, 1, 
                                30, Inf, 0), 
                              ncol = 3, 
                              byrow = TRUE)

sst_rcl_oyster <- classify(sst_mean_c, sst_rcl_mat_oyster)

# Reclassify depth for oysters using classify()
# Create rcl matrix: <-70m = 0, -70m to 0m = 1, >0m = 0
depth_rcl_mat_oyster <- matrix(c(-Inf, -70, 0, 
                                 -70, 0, 1,
                                 0, Inf, 0), 
                               ncol = 3, 
                               byrow = TRUE)

depth_rcl_oyster <- classify(depth_resampled, depth_rcl_mat_oyster)

# Combine suitability layers 
suitable_zones_oyster <- sst_rcl_oyster * depth_rcl_oyster

# Visualize with plot()
plot(suitable_zones_oyster, 
     main = "Oyster Suitability\n (1 = Suitable, 0 = Unsuitable)",
     cex.main = 0.9)
```

## Black Abalone Suitability

### Define Black Abalone Requirements

Black Abalone requires:
- Sea surface temperature: 12.2-18.6°C
- Depth: 0-6 meters below sea level
- Source: SeaLifeBase

```{r species-suitability}
# Reclassify SST for black abalone using classify()
# Suitable: 12.2-18.6°C = 1, Unsuitable: <12.2°C or >18.6°C = 0
sst_rcl_mat_abalone <- matrix(c(-Inf, 12.2, 0,
                                 12.2, 18.6, 1, 
                                 18.6, Inf, 0),  
                               ncol = 3, 
                               byrow = TRUE)

# Apply reclassification matrix to mean SST raster
sst_rcl_abalone <- classify(sst_mean_c, sst_rcl_mat_abalone)

# Reclassify depth for black abalone using classify()
# Suitable: 0-6m below sea level = -6 to 0 in raster values
depth_rcl_mat_abalone <- matrix(c(-Inf, -6, 0,  
                                   -6, 0, 1,     
                                   0, Inf, 0),   
                                 ncol = 3, 
                                 byrow = TRUE)

# Apply reclassification matrix to resampled depth raster
depth_rcl_abalone <- classify(depth_resampled, depth_rcl_mat_abalone)

# Combine suitability layers 
suitable_zones_abalone <- sst_rcl_abalone * depth_rcl_abalone

# Visualize with plot()
plot(suitable_zones_abalone, main = "Black Abalone Suitability\n (1 = Suitable, 0 = Unsuitable)",
     cex.main = 0.9)
```

# Calculate Suitable Area by EEZ

## Prepare EEZ for Area Calculation

```{r prep-eez}
# Transform EEZ to match raster CRS
eez <- st_transform(wc_df, crs(sst_mean_c))

# Rasterize EEZ using rasterize() with rgn_id for field
eez_rast <- rasterize(eez, suitable_zones_oyster, field = "rgn_id")

```

## Calculate Suitable Area for Oysters

```{r calculate-oyster-area}
# Mask suitable areas to EEZ using mask()
eez_suitable_oyster <- mask(suitable_zones_oyster, eez_rast)

# Calculate cell area in km² using cellSize()
cell_area_km2 <- cellSize(eez_suitable_oyster, unit = "km")

# Multiply suitable locations by cell area to get suitable area per cell
suitable_area_oyster <- eez_suitable_oyster * cell_area_km2

# Sum area by EEZ region using zonal()
area_by_region_oyster <- zonal(suitable_area_oyster, eez_rast, fun = "sum", na.rm = TRUE)

# Rename area column for clarity
names(area_by_region_oyster)[2] <- "suitable_area_km2"

# Join results with EEZ sf object to get region names
area_by_region_oyster_joined <- eez %>% 
  left_join(area_by_region_oyster, by = "rgn_id") %>% 
  arrange(desc(suitable_area_km2))
  
oyster_results <- area_by_region_oyster_joined %>% 
  st_drop_geometry() %>% 
  select(rgn, suitable_area_km2) %>%
  head(10)

# Display top 10 regions in table with kable()
kable(oyster_results, 
      caption = "Top 10 EEZ Regions for Oyster Aquaculture",
      col.names = c("Region", "Suitable Area (km²)"),
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r test-area-calc}
# Test that area calculations are reasonable
test_that("Oyster area calculations are valid", {
  # Check total area > 0
  expect_true(sum(area_by_region_oyster$suitable_area_km2, na.rm = TRUE) > 0)
  
  # Check all values >= 0
  expect_true(all(area_by_region_oyster$suitable_area_km2 >= 0, na.rm = TRUE))
  
  # Check that we have results for multiple regions
  expect_true(nrow(area_by_region_oyster) > 0)
})
```

## Map Oyster Suitability by EEZ

```{r map-oyster}
#| fig.width: 10
#| fig.height: 8

# Create map using tmap
tm_shape(area_by_region_oyster_joined) +
  tm_polygons(
    fill = "suitable_area_km2",
    fill.scale = tm_scale_continuous(values = "brewer.yl_gn_bu"),
    fill.legend = tm_legend(title = "Suitable Area (km²)")
  ) +
  tm_borders(col = "gray40", lwd = 0.5) +
  tm_title("Oyster Aquaculture Suitability by EEZ Region") +
  tm_compass(type = "4star", position = c("left", "bottom")) +
  tm_scalebar(position = c("left", "bottom")) +
  tm_layout(legend.outside = TRUE,
            legend.outside.position = "right",
            frame = FALSE)
```

## Calculate Suitable Area for Black Abalone

```{r calculate-species-area}
# Mask suitable areas to EEZ using mask()
eez_suitable_abalone <- mask(suitable_zones_abalone, eez_rast)

# Calculate cell area in km² using cellSize()
cell_area_km2_abalone <- cellSize(eez_suitable_abalone, unit = "km")

# Multiply suitable locations by cell area to get suitable area per cell
suitable_area_abalone <- eez_suitable_abalone * cell_area_km2_abalone

# Sum area by EEZ region using zonal()
area_by_region_abalone <- zonal(suitable_area_abalone, eez_rast, fun = "sum", na.rm = TRUE)

# Rename area column for clarity
names(area_by_region_abalone)[2] <- "suitable_area_km2"

# Join results with EEZ sf object to get region names
area_by_region_abalone_joined <- eez %>% 
  left_join(area_by_region_abalone, by = "rgn_id") %>% 
  arrange(desc(suitable_area_km2))
  
abalone_results <- area_by_region_abalone_joined %>% 
  st_drop_geometry() %>% 
  select(rgn, suitable_area_km2) %>%
  head(10)

# Display top 10 regions in table with kable()
kable(abalone_results, 
      caption = "Top 10 EEZ Regions for Black Abalone Aquaculture",
      col.names = c("Region", "Suitable Area (km²)"),
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```
```{r test-species-area}
# Test that area calculations are reasonable
test_that("Black Abalone area calculations are valid", {
  # Check total area > 0
  expect_true(sum(area_by_region_abalone$suitable_area_km2, na.rm = TRUE) > 0)
  
  # Check all values >= 0
  expect_true(all(area_by_region_abalone$suitable_area_km2 >= 0, na.rm = TRUE))
  
  # Check that we have results for multiple regions
  expect_true(nrow(area_by_region_abalone) > 0)
})
```

## Map Black Abalone Suitability by EEZ

```{r map-species}
#| fig.width: 10
#| fig.height: 8

# Create map using tmap
tm_shape(area_by_region_abalone_joined) +
  tm_polygons(
    fill = "suitable_area_km2",
    fill.scale = tm_scale_continuous(values = "brewer.yl_gn_bu"),
    fill.legend = tm_legend(title = "Suitable Area (km²)")
  ) +
  tm_borders(col = "gray40", lwd = 0.5) +
  tm_title("Black Abalone Aquaculture Suitability by EEZ Region") +
  tm_compass(type = "4star", 
             position = c("left", "bottom")) +
  tm_scalebar(position = c("left", "bottom")) +
  tm_layout(legend.outside = TRUE,
            legend.outside.position = "right",
            frame = FALSE)
```

# Generalizable Function

## Create Suitability Function

```{r create-function}
calculate_suitability <- function(min_temp, max_temp, min_depth, max_depth, species_name) {
  
  # Validate inputs
  if (min_temp >= max_temp) {
    stop("min_temp must be less than max_temp")
  }
  
  if (min_depth >= max_depth) {
    stop("min_depth must be less than max_depth")
  }
  
  if (!is.character(species_name) || species_name == "") {
    stop("species_name must be a non-empty string")
  }
  
  # Create reclassification matrix for sea surface temperature
  sst_rcl_matrix <- matrix(c(-Inf, min_temp, 0,
                             min_temp, max_temp, 1, 
                             max_temp, Inf, 0),
                           ncol = 3,
                           byrow = TRUE)
  # Apply reclassification matrix to mean SST raster
  sst_rcl <- classify(sst_mean_c, sst_rcl_matrix)
  
  
  # Create reclassification matrix for depth
  depth_rcl_matrix <- matrix(c(-Inf, min_depth, 0,
                             min_depth, max_depth, 1,
                             max_depth, Inf, 0),
                           ncol = 3,
                           byrow = TRUE)
  # Apply reclassification matrix to resampled depth raster
  depth_rcl <- classify(depth_resampled, depth_rcl_matrix)
  
  # Combine temperature and depth suitability using multiplication
  suitable_zones <- sst_rcl * depth_rcl
  
  # Mask suitable zones to only include areas within EEZ boundaries
  eez_suitable <- mask(suitable_zones, eez_rast)
  
  # Calculate the area of each raster cell in square kilometers
  cell_area <- cellSize(eez_suitable, unit = "km")
  
  # Multiply suitable locations (1s) by cell area (unsuitable cells with have 0 area)
  suitable_area <- eez_suitable * cell_area
  
  # Sum suitable area within each EEZ region
  area_by_region <- zonal(suitable_area, eez_rast, fun = "sum", na.rm = TRUE)
  
  # Rename the area column to be more descriptive
  names(area_by_region)[2] <- "suitable_area_km2"
  
  # Join area calculations with EEZ spatial data to add region names
  results_joined <- eez %>% 
    left_join(area_by_region, by = "rgn_id") %>% 
    arrange(desc(suitable_area_km2))
  
  # Create a summary table showing top 10 regions
  results_table <- results_joined %>% 
    st_drop_geometry() %>% 
    select(rgn, suitable_area_km2) %>% 
    head(10)
  
  # Create choropleth map showing suitable area by EEZ region
  map <- tm_shape(results_joined) +
    tm_polygons(
      fill = "suitable_area_km2",
      fill.scale = tm_scale_continuous(values = "brewer.yl_gn_bu"),
      fill.legend = tm_legend(title = "Suitable Area (km²)")
    ) +
    tm_borders(col = "gray40", lwd = 0.5) +
    tm_title(paste(species_name, "Aquaculture Suitability by EEZ Region")) +
    tm_compass(type = "4star", position = c("right", "top")) +
    tm_scalebar(stack = "vertical", position = c("left", "bottom")) +
    tm_layout(legend.outside = TRUE,
              legend.outside.position = "right",
              frame = FALSE)
  
  return(list(
    results = results_table,
    full_data = results_joined,
    map = map
  ))
}
```

## Apply Function to Oysters

```{r apply-function-oyster}
#| message: false
#| fig.width: 10
#| fig.height: 8

# Call function with oyster parameters
oyster_output <- calculate_suitability(
  min_temp = 11,
  max_temp = 30,
  min_depth = -70,
  max_depth = 0,
  species_name = "Oyster"
)

# Display results table
kable(oyster_output$results,
      caption = "Top 10 EEZ Regions for Oyster Aquaculture (Function Output)",
      col.names = c("Region", "Suitable Area (km²)"),
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Display map and table
print(oyster_output$map)
```

## Apply Function to Black Abalone
```{r apply-function-species}
#| fig.width: 10
#| fig.height: 8

# Call function with Black Abalone parameters
abalone_output <- calculate_suitability(
  min_temp = 12.2,
  max_temp = 18.6,
  min_depth = -6,
  max_depth = 0,
  species_name = "Black Abalone"
)

# Display map
abalone_output$map

# Display results table
kable(abalone_output$results,
      caption = "Top 10 EEZ Regions for Black Abalone Aquaculture (Function Output)",
      col.names = c("Region", "Suitable Area (km²)"),
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Discussion

## Key Findings

**Oyster Aquaculture**: Central California has the most suitable area (4,069.88 km²), followed by Southern California (3,757.28 km²) and Washington (2,378.31 km²). These three regions contain over 90% of all suitable oyster habitat on the West Coast. Oregon has moderate potential (1,074.27 km²), while Northern California has minimal suitable area.

**Black Abalone Aquaculture**: Total suitable area is dramatically smaller—only 199.88 km² across the entire West Coast. Central California dominates with 152.37 km² (76% of suitable habitat). Washington and Southern California have limited suitable areas, while Oregon and Northern California have none.

**Key Difference**: Black abalone has 20 times less suitable habitat than oysters, potentially due to its narrow temperature tolerance (12.2-18.6°C vs. 11-30°C) and shallow depth requirement (0-6m vs. 0-70m). Both species show peak suitability in Central California, but oysters have viable habitat across all West Coast regions while black abalone is highly restricted.

## Limitations

- **Temporal resolution**: Annual SST averages (2008-2012) may miss critical seasonal temperature extremes or variations important for specific life stages
- **Environmental factors excluded**: Does not account for ocean currents, dissolved oxygen, salinity, wave exposure, water quality, or harmful algal blooms
- **Socioeconomic constraints**: Ignores proximity to markets/infrastructure, conflicts with existing ocean uses, and regulatory/permitting requirements
- **Species requirements**: Based on optimal growth ranges from SeaLifeBase; may not reflect full tolerance ranges or life-stage-specific needs
- **Spatial resolution**: Bathymetry resolution (~500m) may be too coarse to accurately represent ideal depths, especially for shallow-water species like black abalone

# Data Sources

**Sea Surface Temperature:**
- Source: NOAA Coral Reef Watch. NOAA 5km Daily Global Satellite Sea Surface Temperature Anomaly v3.1
- URL: https://coralreefwatch.noaa.gov/product/5km/index_5km_ssta.php

**Bathymetry:**
- Source: GEBCO Compilation Group (2022). GEBCO_2022 Grid
- DOI: 10.5285/e0f0bb80-ab44-2739-e053-6c86abc0289c

**Exclusive Economic Zones:**
- Source: Marine Regions
- URL: https://www.marineregions.org/

**Species Requirements:**
- Source: SeaLifeBase
- URL: https://www.sealifebase.org/